---
title: "Inferring a trajectory with slingshot"
output: github_document
bibliography: bibliography.bib
---
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(velocyto.R)
library(Seurat)
library(SeuratWrappers)
```

```{r}
seu_trajectory <- read_rds("data/adipose_differentiation-merick/seu_trajectory.rds")
feature_info <- read_tsv("data/adipose_differentiation-merick/feature_info.tsv")
feature_mapper <- function(x) {feature_info %>% dplyr::slice(base::match(symbol, x)) %>% dplyr::pull(feature_id)}
```

## Estimating the future state of a cell

Our dataset contains two assays: the number of spliced and unspliced reads.

```{r}
seu_trajectory@assays$spliced
seu_trajectory@assays$unspliced
```

Let's get a feeling of how this data actually looks like. Let's take the gene that was most associated with pseudotime according to tradeSeq.

```{r}
tibble(
  spliced = seu_trajectory@assays$spliced@counts["ENSMUSG00000030116.14", ],
  unspliced = seu_trajectory@assays$unspliced@counts["ENSMUSG00000030116.14", ]
) %>% 
  ggplot(aes(spliced, unspliced)) +
    geom_jitter()
```

From this, we want to predict a future state of the cell.

A naive approach would be to use the `unspliced` matrix as a prediction of what the `spliced` matrix will look like in the future. However, such an approach would make some big assumptions, such as identical splicing rates and degradation rates for each gene, which are easily violated [@mannoRNAVelocitySingle2018].

Instead, we need to have an idea on the balance between the generation of new spliced mRNA (i.e. splicing rates) and removal of spliced mRNA (i.e. degradation rates). We don't need to know the exact speed of either of these processes, but rather the ratio between the two. If we would know this, we would be able to use the current amount of unspliced mRNA and see how much is generated _in relationship_ to the degradation of old spliced mRNA. We can only do this for a short time interval where we can assume that the unspliced mRNA is in a steady state, and does not change too much based on gene regulation (or post-transcriptional regulation).

How can we estimate the ratio of splicing vs degradation, i.e. γ? We might assume that if the number of mRNAs is very high or very low, it has reached a steady state. In this steady state, the ratio between splicing and degradation is simply equal to the ratio between unspliced and spliced mRNAs. This way of estimating the γ is implemented in the velocyto.R package. Note however that this makes many assumptions, which could be easily violated in real settings:

- It assumes that there is one common γ for each gene. Although some data suggests this might be ok for most cell types [@mannoRNAVelocitySingle2018], some more in depth study across cell types might be necessary to really prove this assumption.
- It assumes that the gene reaches a steady state at the lowest and highest expression levels within the dataset.

As this is still a topic of active research, some alternative models are also being proposed, e.g. the ones created in the python scvelo package (https://scvelo.readthedocs.io/about.html). These do not make less assumptions, but at the cost of being more computationally intensive. Some benchmarking will undoubtedly be necessary to assess whether these alternatives do generate more accurate models.

Another important point to consider is that the observed counts of both unspliced and spliced reads are very noisy. Estimating the future state of a cell based on these values would therefore amplify this noise even further, leading to highly variable and inaccurate estimations. One way to mitigate this is by pooling the unspliced and spliced counts from similar cells (similarly to imputation methods) through a kNN. Keep in mind that this introduces yet another bias into the analysis, and that subtle differences of just a couple of cells are possibly "smoothed out" by this kNN.

The estimation of labmda and projection to a future state are both implemented in the `velocyto.R::gene.relative.velocity.estimates()` function. The SeuratWrappers package provides a helpful wrapper for this to run this within Seurat:

```{r}
seu_trajectory <- RunVelocity(seu_trajectory, spliced.average = 0.2, unspliced.average = 0.05)
```

Because of the computational complexity of estimating the lambda, we add some extra filtering on the spliced and unspliced counts. I know, this introduces yet another bias, so feel free to loosen this filtering if you have the patience.

We now have extra information in the `seu_trajectory@tools$RunVelocity`. Among which are the projected (future) state of the cell (this is stored as a non-sparse matrix, another reason why some prior filtering is important):

```{r}
seu_trajectory@tools$RunVelocity$projected[1:10, 1:10]
```

The predicted gamma values for each gene
```{r}
head(seu_trajectory@tools$RunVelocity$gamma)
qplot(seu_trajectory@tools$RunVelocity$gamma) + scale_x_log10() + theme_classic()
```

And the kNN if you would want to use it for some other purpose:

```{r}
seu_trajectory@tools$RunVelocity$cellKNN[1:10, 1:10]
```

To understand this better, let's plot these values for the Dpp4 gene (the `show.gene` argument here will estimate the γ for this gene only instead of all):


```{r, fig.width=14, fig.height=5}
gene.relative.velocity.estimates(
  GetAssayData(seu_trajectory, slot = "data", assay = "spliced"),
  GetAssayData(seu_trajectory, slot = "data", assay = "unspliced"),
  cell.emb = Embeddings(seu_trajectory, "umap"),
  show.gene = "ENSMUSG00000030116.14",
  old.fit = Tool(seu_trajectory, slot = "RunVelocity")
)
```

These plots show the sliced counts, unspliced counts, the relationship between the two and an estimate of whether the gene's expression is decreasing (blue) or increasing (red). The estimate for lamba is shown in the third plot as a red dashed line. As you can see, most cells are under this line, meaning that their expression is decreasing. Also note the red dots to the right: although this suggests that some cells have a high upregulation of this gene, a more plausible explanation would be that these cells had relatively high unspliced reads just by chance. What we're probably seeing is something akin to 'overinflation of fold-changes' just like in bulk data.

Investigating the velocity of one feature can be misleading, because wrong annotations, problems with the estimation of γ and low counts. More useful  (or in the very least robust) is therefore to pool the information across the genes, and see the direction into which the cells are moving.

To do this, we have two options: (1) we can use an existing dimensionality reduction and map the projected cells onto it, or (2) we can either plot both the cells and their projected counterparts on a new dimesionality reduction.

The first option is the one that you're most probably want to use, given that you often already have a good dimensionality reduction that you use across the project. But at the same time, it is the most difficult one: 

- If the dimensionality reduction allows you to map new data, this is pretty straightforward.
  We can just map the projected state of the cell in the embedding.
  PCA is the best known example of such a method.
  
- If the dimensionality reduction does not have a mapping function, we have to use the existing points in the embedding to estimate where the projected cells will lie.
  This includes most dimensionality reduction methods, such as MDS, t-SNE and UMAP.
  
Let's try out the second approach, with the following dimensionality reductions:


```{r, message=FALSE, warning=FALSE}
# for demonstration we use a subsampled seurat object here because the following steps are very time consuming
set.seed(1)
seu_velocity <- seu_trajectory[, sample(colnames(seu_trajectory), 2000)]


# MDS, we use the landmark mds because it is much faster and memory efficient
set.seed(1)
lmds <- dyndimred::dimred_landmark_mds(Matrix::t(seu_velocity@assays$spliced@scale.data))
colnames(lmds) <- paste0("lmds_", seq_len(ncol(lmds)))
lmds_object <- CreateDimReducObject(lmds, key = "lmds_", assay = "spliced")
seu_velocity@reductions$lmds <- lmds_object

# UMAP, based on the R uwot implementation
set.seed(1)
seu_velocity <- RunUMAP(seu_velocity, dims = 1:50, umap.method = "uwot")
```

```{r}
cell_pal <- function(cell_cats, pal_fun) {
  categories <- sort(unique(cell_cats))
  pal <- setNames(pal_fun(length(categories)), categories)
  pal[cell_cats]
}
cell_colors <- cell_pal(seu_velocity$seurat_clusters, scales::brewer_pal("qual", "Set2"))
cell_colors_clust <- cell_pal(seu_velocity$seurat_clusters, scales::hue_pal())
names(cell_colors) <- names(cell_colors_clust) <- colnames(seu_velocity)

cc_mds <- velocyto.R::show.velocity.on.embedding.cor(
  emb = Embeddings(seu_velocity, "lmds"),
  vel = Tool(seu_velocity, slot = "RunVelocity"),
  show.grid.flow = TRUE,
  cell.colors = cell_colors
)
```

We have to tune some parameters here to make this chaotic plot more interpretable

```{r}
cc_mds <- velocyto.R::show.velocity.on.embedding.cor(
  emb = Embeddings(seu_velocity, "lmds"),
  vel = Tool(seu_velocity, slot = "RunVelocity"),
  show.grid.flow = TRUE,
  cell.colors = cell_colors,
  
  # algorithm parameters
  grid.n = 15, # number of points to plot for the grid
  n = 200, # smooth more over nearby cells
  min.grid.cell.mass = 50, # higher values will mean only arrows in highly dense regions
  
  # aesthetic paramters
  cex = 0.5,
  cell.border.alpha = 0,
  arrow.scale = 1,
  arrow.lwd = 1.5,
)
```

And on UMAP

```{r}
cc_mds <- velocyto.R::show.velocity.on.embedding.cor(
  emb = Embeddings(seu_velocity, "umap"),
  vel = Tool(seu_velocity, slot = "RunVelocity"),
  show.grid.flow = TRUE,
  cell.colors = cell_colors,
  
  # algorithm parameters
  grid.n = 15,
  n = 200,
  min.grid.cell.mass = 1,
  
  # aesthetic paramters
  cex = 0.5,
  cell.border.alpha = 0,
  arrow.scale = 6,
  arrow.lwd = 1.5,
)
```

Plotting RNA velocity vectors is still a bit of a tricky business, because there are many factors to consider:
 - The vectors are inherintly very noisy, so you need to average over many cells
 - You can't average too much, otherwise you could miss subtle local effects
 - High density regions highly influence the vectors of nearby low density regions (a similar problem exists with imputation methods)

Observe how even a small change in parameters (mds vs umap) impacts the velocity vectors. Although the main message remains the same
This result will likely become more robust if we include more cells, instead of the subsample of cells that was used here.

Plotting RNA velocity on an existing dimensionality reduction may introduce some artifacts, particularly for cells at the "border". What if their future cell state is not contained in any of the other cells in the dataset? Observe how the arrows at the end of the two terminal populations often point inwards. This makes sense, given that the 

Alternatively, we can also create an embedding in wich both the cells and their projected state are embedded at the same time. velocyto.R only provides functions to do this for PCA and UMAP, e.g.:

```{r}
cc_mds <- velocyto.R::tSNE.velocity.plot(
  vel = Tool(seu_velocity, slot = "RunVelocity"),
  show.grid.flow = TRUE,
  cell.colors = cell_colors,
  
  # algorithm parameters
  grid.n = 15,
  min.grid.cell.mass = 1,
  
  # aesthetic paramters
  cex = 0.5,
  cell.border.alpha = 0,
  # arrow.scale = 6,
  arrow.lwd = 1.5,
)
```

This may take quite a while to run, given that 